Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2019-06-05 review/response.Rmd

**Mention holes/cmass with reference in Sec. 2.2, but keep rescaling information in the appendix?**

8) Check the spelling of words for clarity and consistency, e.g, splines2D/splines2d; convex/1-convex, MIC/MICe, and so on.

**We have made the spelling consistent where referring to the same index, and provided explanations where the different spelling indicates differences in index calculation.**

Reviewer #2:
Summary of the contribution:
The authors present a simulation study to evaluate various projection pursuit indexes for detecting specific bivariate patterns in the data. Moreover, two data sets from physics are examined using PPI.

Major remarks
The simulation study targets three particular structures the pipe, the sine, and the spiral. Some motivation and justification for selecting these structures should be added.

 **We have added a statement in Section 3.1.1 outlining the motivation for the three structures considered in the simulation study.**

Second, the simulation study uses six-dimensional data of which four dimensions con- tribute nuisance and the other two the intended structure. Implicitly, this is clear, but the description does not make this explicit and uses the general notion of p dimensions instead. Please make this more explicit and also provide some justification that this set-up is adequate for targeting the problem at hand.

**We have added an explicit statement that we fix p=6. As mentioned in the statement p=6 results in two independent planes showing the nuisance distribution and one plane showing the structured distribution. This set-up is adequat for most of our study where we follow interpolated paths either between nuisance only projections or from nuisance views to structured projections.**

Thirdly, the simulations refer to sequences of projections (the horizontal axis in plots in Figures 2, 3, and 4. As I understand these sequences are a consequence of optimising the corresponding projection pursuit index. How can you then actually control the two different scenarios referred to in section 3.2 and 3.3? Or, do you extract a specified sequence of projections? Moreover, how do you make sure that you have the same sequence length for each trial? Please clarify!

**Figures 2 and 3 present the trace for a specified sequence of projections (since in the simulation study the nuisance and structured views are known). This is explained in Section 3.1.2, and we have added a sentence clarifying this in Section 3.2 and made the wording more explicit in Section 3.3. Figure 4 is different as in this case the path is indeed the result of optimising an index. However, we only optimise the TIC index highlighted in red, while the other index values are shown for comparison along the optimisation path.**

**(In my opinion all of this was already very clear before. When discussing this before you suggested making the Figure caption more clear, but it seems they are already very clear and I would not want to make them longer.)**

The optimisation methods are further discussed in section 3.4 and depend on the struc- ture present in the data. What is the value of these investigations for more realistic settings in which the structure is not known.

**(I would list: demonstrate potential as PPI, understand shortcomings, and based on that develop strategies, e.g. the two-step optimisation. Add something like this in Section 3.4?)**

Finally, the relationship between the simulation study and the examples using the astrophysics data remains unclear. How are the physics applications linked to the simulation study and vice versa? Currently, the two aspects seem to me rather independent and could actually be presented in separate manuscripts. Moreover, since some rather important aspects of the simulation study have been put into an appendix. For the COST audience certainly the details of the simulation study are more relevant than the astrophysics application.

**(Point to initial motiviation, what else should we say here?)**

Minor remarks
When introducing the various PPIs some terms and notations are used which are not properly introduced, e.g. alpha hull, convex. Check which ones actually need to be properly defined as they are needed for further steps and for which ones you could just refer to the literature avoiding unnecessary details.

**(Still need to do this)**

In Figure 3, each plot shows a vertical blue line. Please explain what this line represents.

**We have added such a sentence to the figure caption.**

p. 14, l. 26: optimist should be optimise

**We have fixed the typo.**

### 2019-06-05 scagnostics.Rmd
Adding comments from reviewer 2.

Three families of data simulations are used for examining the behaviour of the index functions. Each generates  structure in two variables, with the remaining variables containing various types of noise. This is a very simple construction, because there is no need for projection pursuit to find the structure, one could simply use the PPIs on pairs of variables. However, it serves the purpose to also evaluate the PPIs. The three data families are explained below. In each set, $n$ is used for the number of points, $p$ is the number of dimensions, and $d=2$ is the projection dimension.

@@

The structured projection is in variables 5 and 6 ($x_5, x_6$). Two samples sizes are used: $n=(100, 1000)$. All variables are standardized to have mean 0 and standard deviation 1. Figure \ref{fig:data} shows samples from each of the families, of the nuisance and structured pairs of variables. Table \ref{tab:indexTable} compares the PPIs for structured projections against those for nuisance variables, based on 100 simulated data sets of each type, using sample size 1000. The lower and upper show the 5th and 95th percentile of values. The holes index is sensitive only to the pipe distribution. All other indexes, except convex show distinctly higher values for the structured projections. The convex index shows the inverse scale to other indices, thus (1-convex) will be used in the assessment of performance of PPIs. The scale for the holes index in its original implementation is smaller than the others ranging from about 0.7 through 1, so it is re-scaled in the performance assessment so that all indices can be plotted on a common scale of 0-1 (details in Appendix). Similarly, the TIC index is re-scaled depending on sample size.

@@

Figure \ref{fig:plotV1V2toV3V4} shows the traces representing the index values calculated across a tour between a pair of nuisance projections. The range of each axis is set to be the limits of the index, as might be expected over many different data sets, 0 to 1. Each projection in the interpolation will also be noise. Two different sample sizes are show, $n=100$ as a dashed line, and $n=1000$ as a solid line. The ideal trace is a smooth function, with relatively low values, and no difference between the sample sizes. A major feature to notice is that the scagnostics produce noisy functions, which is problematic, because small changes in the projection result in big jumps in the value. This will make them difficult to optimize. On the other hand holes, dcor2d, splines2d, MIC and TIC are relatively smooth functions.

@@

Figure \ref{fig:wIntermediate} shows the PPIs for a tour sequence between a nuisance and structured projection. A long sequence is generated where the path runs from $x_1$-$x_2$, $x_1$-$x_5$, $x_5$-$x_6$, in order to see some of the intricacies of holes index. Sample size is indicated by line type: dashed being $n=100$ and solid is $n=1000$. The beginning of the sequence is the nuisance projection and the end is the structured projection. The index values for most PPIs increases substantially nearing the structured projection, indicating that they "see" the structure. Some indexes see all three structures: scagnostics, MIC and TIC,  which means that they are flexible indexes capable of detecting a range of structure. [@Grimm2016]'s indexes, dcor2d and splines2d, are excellent for detecting the sine, and they can see it from far away, indicated by the long slow increase in index value. The holes index easily detects the pipe, and can see it from a distance, but also has local maxima along the tour path. The scagnostic index, stringy, can see the structure but is myopic, only when it is very close. Interestingly the scagnostic, skinny, sees the spiral from a distance.

@@

```{r wIntermediate, fig.width=6, fig.height=7, fig.cap="PPIs for projections along an interpolation between nuisance and structured projections, following $x_1$-$x_2$ to $x_1$-$x_5$ to $x_5$-$x_6$ (solid lines: $n=1000$; dashed: $n=100$). Peaks at the end of the sequence indicate the index sees the structure. The scagnostics, MIC and TIC see all three structures, so are more flexible for general pattern detection. Holes only responds to the pipe, and is a multimodal function for this data with a local maximum at $x_1$-$x_5$. "}
intermediatePlane <- which(attributes(interpolate(t3))$new_basis)[2]
namesFixed <- str_replace_all(fullResPlanned$name, "sin", "sine")
fullResMeltPlanned <- fullResPlanned %>%
  mutate(name = namesFixed) %>%
  mutate(convex = 1-convex) %>%
  rename("1-convex"=convex) %>%
  rowwise() %>%
  mutate(holes = rescaleHoles(holes)) %>%
  gather(PPI, value, -t, -name, -size) %>%
  filter(PPI != "cmass") %>%
  mutate(PPI = fct_relevel(PPI, "holes", "1-convex", "skinny", "stringy", "dcor2d", "splines2d", "MIC", "TIC"))
fullResMeltPlanned %>% 
  ggplot(aes(x=t, y=value)) + 
  geom_line(aes(group=size, linetype=factor(size, levels=c("1000", "100")))) + 
  ylim(c(0,1)) +
  geom_vline(xintercept=intermediatePlane, color="blue", alpha=0.3) +
  facet_grid(PPI~name) + 
  theme(legend.position="none") + xlab("Sequence of projections (t)") + ylab("PPI value")
```

@@

The optimization is done in two stages, a scouting phase using method 2, and a refinement stage using method 3.  For the scouting we use $\alpha = 0.5$ and stopping condition of maximum 5000 tries, and we optimist the TIC index.

>>>>

Three families of data simulations are used for examining the behaviour of the index functions. Each generates  structure in two variables, with the remaining variables containing various types of noise. This is a very simple construction, because there is no need for projection pursuit to find the structure, one could simply use the PPIs on pairs of variables. However, it serves the purpose to also evaluate the PPIs. The three data families are explained below. In each set, $n$ is used for the number of points, $p$ is the number of dimensions, and $d=2$ is the projection dimension. 
\textcolor{red}{The three structures were selected to cover both functional and non-functional dependence, different types of nuissance distributions and different structure size and squintability properties.}

@@

\textcolor{red}{In the following we fix $p=6$ (this means that we have two independent planes containing the nuisance distribution, and one plane containing the structured distribution), and}
the structured projection is in variables 5 and 6 ($x_5, x_6$). Two samples sizes are used: $n=(100, 1000)$. All variables are standardized to have mean 0 and standard deviation 1. Figure \ref{fig:data} shows samples from each of the families, of the nuisance and structured pairs of variables. Table \ref{tab:indexTable} compares the PPIs for structured projections against those for nuisance variables, based on 100 simulated data sets of each type, using sample size 1000. The lower and upper show the 5th and 95th percentile of values. The holes index is sensitive only to the pipe distribution. All other indexes, except convex show distinctly higher values for the structured projections. The convex index shows the inverse scale to other indices, thus (1-convex) will be used in the assessment of performance of PPIs. The scale for the holes index in its original implementation is smaller than the others ranging from about 0.7 through 1, so it is re-scaled in the performance assessment so that all indices can be plotted on a common scale of 0-1 (details in Appendix). Similarly, the TIC index is re-scaled depending on sample size.

@@

Figure \ref{fig:plotV1V2toV3V4} shows the traces representing the index values calculated across a tour between a pair of nuisance projections.
\textcolor{red}{The tour path is generating by interpolating between the two independent nuisance planes, i.e. from the projection onto $x_1$-$x_2$ to one onto $x_3$-$x_4$.}
The range of each axis is set to be the limits of the index, as might be expected over many different data sets, 0 to 1. Each projection in the interpolation will also be noise. Two different sample sizes are show, $n=100$ as a dashed line, and $n=1000$ as a solid line. The ideal trace is a smooth function, with relatively low values, and no difference between the sample sizes. A major feature to notice is that the scagnostics produce noisy functions, which is problematic, because small changes in the projection result in big jumps in the value. This will make them difficult to optimize. On the other hand holes, dcor2d, splines2d, MIC and TIC are relatively smooth functions.

@@

Figure \ref{fig:wIntermediate} shows the PPIs for a tour sequence between a nuisance and structured projection. A long sequence is generated where the path 
\textcolor{red}{interpolates between projections onto}
$x_1$-$x_2$, $x_1$-$x_5$, $x_5$-$x_6$, in order to see some of the intricacies of holes index. Sample size is indicated by line type: dashed being $n=100$ and solid is $n=1000$. The beginning of the sequence is the nuisance projection and the end is the structured projection. The index values for most PPIs increases substantially nearing the structured projection, indicating that they "see" the structure. Some indexes see all three structures: scagnostics, MIC and TIC,  which means that they are flexible indexes capable of detecting a range of structure. [@Grimm2016]'s indexes, dcor2d and splines2d, are excellent for detecting the sine, and they can see it from far away, indicated by the long slow increase in index value. The holes index easily detects the pipe, and can see it from a distance, but also has local maxima along the tour path. The scagnostic index, stringy, can see the structure but is myopic, only when it is very close. Interestingly the scagnostic, skinny, sees the spiral from a distance.

@@

```{r wIntermediate, fig.width=6, fig.height=7, fig.cap="PPIs for projections along an interpolation between nuisance and structured projections, following $x_1$-$x_2$ to $x_1$-$x_5$ to $x_5$-$x_6$ (solid lines: $n=1000$; dashed: $n=100$). The vertical blue line indicates the position of the projection onto $x_1$-$x_5$ in the sequence. Peaks at the end of the sequence indicate the index sees the structure. The scagnostics, MIC and TIC see all three structures, so are more flexible for general pattern detection. Holes only responds to the pipe, and is a multimodal function for this data with a local maximum at $x_1$-$x_5$. "}
intermediatePlane <- which(attributes(interpolate(t3))$new_basis)[2]
namesFixed <- str_replace_all(fullResPlanned$name, "sin", "sine")
fullResMeltPlanned <- fullResPlanned %>%
  mutate(name = namesFixed) %>%
  mutate(convex = 1-convex) %>%
  rename("1-convex"=convex) %>%
  rowwise() %>%
  mutate(holes = rescaleHoles(holes)) %>%
  gather(PPI, value, -t, -name, -size) %>%
  filter(PPI != "cmass") %>%
  mutate(PPI = fct_relevel(PPI, "holes", "1-convex", "skinny", "stringy", "dcor2d", "splines2d", "MIC", "TIC"))
fullResMeltPlanned %>% 
  ggplot(aes(x=t, y=value)) + 
  geom_line(aes(group=size, linetype=factor(size, levels=c("1000", "100")))) + 
  ylim(c(0,1)) +
  geom_vline(xintercept=intermediatePlane, color="blue", alpha=0.3) +
  facet_grid(PPI~name) + 
  theme(legend.position="none") + xlab("Sequence of projections (t)") + ylab("PPI value")
```

@@

The optimization is done in two stages, a scouting phase using method 2, and a refinement stage using method 3.  For the scouting we use $\alpha = 0.5$ and stopping condition of maximum 5000 tries, and we optimised the TIC index.
